---
title: "R Markdown Advanced Example"
author: "Damian Pavlyshyn"
date: "4/22/2020"
output:
  html_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "D:/Users/Damian/stats32")
```

# Chunk options

Sometimes we may want not include the code or outpout of a particular chunk.
We can do this using "chunk options" that we include in the `{r <name> <options>}` part of the chunk opening.

`echo=FALSE` will suppress code from the output, but will display the output.
This is useful, for example, when you have a complicated plotting specification that isn't relevant to the analysis itself.
However, **please do not use this option in your final project!** I want to see all your code!
```{r name, echo=FALSE}
1 + 1
```

We can also suppress both the code and the output with the `include=FALSE` option.
In the following invisivle chunk, I set the variable `v`.
```{r include=FALSE}
v <- c(1, 2, 3)
```

Now I can access the variable even though the code that set it doesn't appear in the output file.
```{r}
v
```

Often when you run R code, it prints a bunch of information in addition to the actual output.
By default, this is included in the output document, but it can be suppressed using `message=FALSE`.
```{r}
library(tidyverse)
scores <- read_csv("example.csv")
scores
```

The last very useful options are the `fig.height` and `fig.width` options, wihch control the size of any output figures.
```{r fig.height=5, fig.width=2}
ggplot(scores, aes(x = Year)) + 
  geom_bar()
```

## The setup chunk

We can set options for all chunks in the document with a sepcial "setup" chunk at the beginning of the document.
This document's setup chunk also contains a line that sets the root directory, which is preferrable to using `setwd()` for some technical reasons.

# Header options

We can specify a lot of interesting formatting options in the header, as demonstrated in this document's header.
In this case, we've numbered the sections and added an automatic table of contents.

A good summary of available and exotic header options can be found [here](https://bookdown.org/yihui/rmarkdown/html-document.html).
